---
title: "Introduction to GIS in R"
author: "Samuel Bosch and Jorge Assis"
date: "13/03/2018"
output:
  ioslides_presentation:
    css: slides.css
    smaller: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

```{r install, include=FALSE}
pkgs <- c('dplyr', 'openxlsx', 'raster', 'readr', 'rgbif',  'robis', 'sdmpredictors', 'sf', 'tidyr')
for(pkg in pkgs) {
  if(!requireNamespace(pkg)) {
    install.packages(pkg, repos = 'https://cloud.r-project.org')
  }
}
```

```{r demodata, include=FALSE}
if(!file.exists('data/Abra_alba_2000.csv')) {
  abra <- robis::occurrence("Abra alba", year=2000)
  readr::write_csv(abra, 'data/Abra_alba_2000.csv')
  readr::write_delim(abra, 'data/Abra_alba_2000.txt', delim = '\t')
  openxlsx::write.xlsx(abra, 'data/Abra_alba_2000.xlsx', sheetName = 'Abra alba', row.names = FALSE)
}
```


## Outline

Goal: give an overview of the most important packages to support you with preparing and exploring data for SDM and for presenting your results

- Importing data
- Getting data
- Manipulating data
- Visualizing data

# Importing data

## Text delimited
  
```{r readr, warning=FALSE, message=FALSE, cache=TRUE}
library(readr)

# comma separated
abra <- read_csv('data/Abra_alba_2000.csv')

# tab separated
abra <- read_delim('data/Abra_alba_2000.txt', delim = '\t')
```
  
## Excel: openxlsx

```{r openxlsx, message=FALSE, cache=TRUE}
# install.packages('openxlsx')
library(openxlsx)

# First sheet
abra <- read.xlsx('data/Abra_alba_2000.xlsx', 1)

# Sheet by name
abra <- read.xlsx('data/Abra_alba_2000.xlsx', sheet = 'Abra alba')
```

## Zipfile

```{r zip, cache=TRUE}
# Download a zip file
temp <- tempfile('nsbs', fileext = '.zip')
download.file('http://ipt.vliz.be/eurobis/archive.do?r=nsbs&v=1.1', 
              temp)

# Read occurence.txt file from the zip file
data <- read_delim(unz(temp, 'occurrence.txt'), delim = '\t')
``` 

## Shapefile

```{r shp, message=FALSE, warning=FALSE, cache=TRUE}
library(sf)
meow <- read_sf('data/MEOW/meow_ecos.shp')

library(raster)
meow <- shapefile('data/MEOW/meow_ecos.shp')
```

## Raster

```{r raster, cache=TRUE}
library(raster)

sst <- raster('data/BO2_tempmean_ss_lonlat.tif')
sst
```

# Getting data

## Environmental

```{r environmental, message=FALSE, warning=FALSE, cache=TRUE}
library(sdmpredictors)

# exploring the marine datasets
datasets <- list_datasets(terrestrial = FALSE, marine = TRUE)
View(datasets)

# exploring the layers
layers <- list_layers(datasets)
View(layers)

# download specific layers to the data directory
environment <- load_layers(c('BO2_tempmean_ss', 'BO_bathymean'), 
                           datadir = 'data')
```

## Occurrences: OBIS

```{r robis, message=FALSE, cache=TRUE}
library(robis)
ptevol <- occurrence('Pterois volitans') # red lionfish
colnames(ptevol)
```

## Filter by year

```{r robis_year, cache=TRUE}
occurrence('Pterois volitans', year = 1991, 
           fields = c('id', 'decimalLongitude', 'decimalLatitude', 'year'))
```

## Filter by date

```{r robis_data, cache=TRUE}
occurrence('Pterois volitans', startdate = '1980-03-01', enddate = '1980-10-30', 
           fields = c('id', 'decimalLongitude', 'decimalLatitude', 'eventDate'))
```

## Filter by depth

```{r robis_depth, cache=TRUE}
occurrence('Pterois volitans', startdepth = 100, enddepth = 200,
           fields = c('id', 'decimalLongitude', 'decimalLatitude', 'depth'))
```

## Filter by polygon

With <http://iobis.org/maptool>) we can create a WKT for the Perhentian and Redang Islands and use this to query all occurrences.

```{r robis_geom, cache=TRUE}
islands <- occurrence(geometry = 
      paste('POLYGON ((102.68921 6.05862','102.57111 5.95346',
                      '103.07785 5.49980','103.25226 5.62555',
                      '103.07648 5.87970','102.68921 6.05862))',sep=', '))
```

## Filter by dataset

```{r robis_dataset, cache=TRUE}
datasets <- dataset(q='Adriatic-INTERREG')
datasets$name
```

## Filter by dataset

```{r robis_dataset2, cache=TRUE}
occ <- occurrence(resourceid = datasets$id, year = 1998,
           fields = c('id', 'decimalLongitude', 'decimalLatitude', 'resourceID'))
head(occ)
```

## Filter by quality control

Vandepitte L., Bosch S., Tyberghein L., Waumans F., Vanhoorne B., Hernandez F., De Clerck O., & Mees J. (2015) **Fishing for data and sorting the catch: assessing the data quality, completeness and fitness for use of data in marine biogeographic databases**. Database, 2015. <http://dx.doi.org/10.1093/database/bau125>

```{r helpqc, eval=FALSE}
?qcflags
```

Categories of quality control flags:

- Data format
- Taxonomy
- Completeness
- Geography
- Depth
- Outliers

## Some filters: quality control

- 3\. Taxonomy: is the taxon level genus or lower?
- 4\. Geography (lat/lon): are the latitude/longitude values different from zero?
- 5\. Geography (lat/lon): are the latitude/longitude values within their possible boundaries? (world coordinates)
- 7\. Completeness (date/time): is the sampling year (start/end) completed and valid?

```{r robis_qc, cache=TRUE}
qc3457 <- occurrence(resourceid = datasets$id, qc = c(3,4,5,7), year = 1998)
```

## Occurrences: GBIF

```{r rgbif, cache=TRUE, warning=FALSE}
library(rgbif)

occ_search(scientificName = 'Pterois volitans', limit = 10)
```

# Manipulating data

## Filtering

```{r dplyr_filter, eval=FALSE}
library(dplyr)

data <- occurrence('Sargassum')
data %>% filter(scientificName == 'Sargassum muticum' & yearcollected > 2005)
```

## Reordering

```{r dplyr_arrange, eval=FALSE}
data %>% arrange(datasetName, desc(eventDate)) %>% print(n=10)
```

## Selecting and renaming columns

```{r dplyr_select, eval=FALSE}
data %>% select(scientificName, eventDate, lon=decimalLongitude, lat=decimalLatitude) %>% print(n=10)
```

## Unique combinations

select() can be used with distinct() to find unique combinations of values:

```{r dplyr_distinct, eval=FALSE}
data %>% select(scientificName, locality) %>% distinct()
```

## Adding columns

```{r dplyr_addcol, eval=FALSE}
data %>% tbl_df %>% mutate(zone = .bincode(minimumDepthInMeters, breaks=c(0, 20, 100))) %>% 
  select(minimumDepthInMeters, zone) %>% 
  filter(!is.na(zone)) %>% 
  print(n = 10)
```

## Aggregation

```{r dplyr_summarise, eval=FALSE}
data %>% summarise(lat_mean = mean(decimalLatitude), lat_sd = sd(decimalLatitude))
data %>% group_by(scientificName) %>% summarise(records=n(), datasets=n_distinct(datasetName))
```

## Extract data from raster

```{r raster_extract, cache=TRUE, message=FALSE}
library(raster)
environment <- sdmpredictors::load_layers(c('BO2_tempmean_ss', 'BO_bathymean'), 
                                          datadir = 'data')
abra <- robis::occurrence("Abra alba", year=2000,
                          fields = c('id', 'decimalLongitude', 'decimalLatitude', 'qc', 'depth'))

extract(environment, abra[,c('decimalLongitude', 'decimalLatitude'),drop=FALSE])
```

## Overlap with polygons

```{r sf_intersection, cache=TRUE}
library(sf)
abra_points <- st_as_sf(abra, coords = c('decimalLongitude', 'decimalLatitude'), crs = 4326)
meow <- read_sf('data/MEOW/meow_ecos.shp')

abra_meow <- st_intersection(abra_points, meow)
colnames(abra_meow)
```

## Spatial thinning

```{r, eval=FALSE}
if (!require('devtools')) {
	install.packages('devtools', repo='http://cran.rstudio.com', dep=TRUE)
}
devtools:::install_github('mlammens/spThin')
```

## Spatial thinning

```{r spthin, cache=TRUE, message=FALSE}
library(spThin)

abra_thin <- spThin(abra, x.col = 'decimalLongitude', y.col = 'decimalLatitude',
                    dist = 10, great.circle.distance = TRUE)
NROW(abra)
NROW(abra[abra_thin@samples[[1]],,drop=FALSE])
```

## Create coastline raster

```{r coastmask, cache=TRUE}
coast_mask <- function(layer) {
  edges <- raster::boundaries(layer, type="inner")
  values <- getValues(edges)
  is.na(values) | values == 0
}
l <- raster('data/BO2_tempmean_ss_lonlat.tif')
mask <- coast_mask(l)
l[mask] <- NA
```

# Visualizing data

## ggplot2

<http://ggplot2.tidyverse.org/reference/>

All ggplot2 plots with a call to `ggplot()`, supplying default data and aesthethic mappings, specified by `aes()`. You then add layers, scales, coords and facets with `+`. To save a plot to disk, use `ggsave()`.

## Preparing data

```{r abraplotprep, cache=TRUE}
library(dplyr)
envabra <- extract(environment, abra[,c('decimalLongitude', 'decimalLatitude')])
data <- cbind(abra, envabra) %>% 
  mutate(ObservationDepth = -1*depth, Bathymetry = BO_bathymean) %>%
  select(ObservationDepth, Bathymetry)
data <- data[complete.cases(data),,drop=FALSE]
data_long <- tidyr::gather(data)
```

## Points

```{r abrapoints, cache=TRUE}
library(ggplot2)
ggplot(data, aes(x=Bathymetry, y=ObservationDepth)) +
  geom_point()
```

## Boxplots

```{r abrabox, cache=TRUE}
ggplot(data_long, aes(x=key, y=value)) + 
  geom_boxplot()
```

## Histogram

```{r abrahist, cache=TRUE}
ggplot(data_long, aes(x=value)) +
  geom_histogram(binwidth = 5, alpha = 0.5) +
  facet_grid(key ~ .)
```

## Frequency polygons

```{r abrafreqpoly, cache=TRUE}
ggplot(data_long, aes(x=value, colour=key)) +
  geom_freqpoly(binwidth = 5)
```

## Density

```{r abradensity, cache=TRUE}
ggplot(data_long, aes(x=value, fill=key, colour=key)) +
  geom_density(alpha=0.5)
```

## Correlation between rasters

```{r rastercor, cache=TRUE}
library(sdmpredictors)
layers <- list_layers('Bio-ORACLE', version = 2)
meanss <- layers %>% filter(is_surface & grepl('mean[_]', layer_code))
corr <- layers_correlation(meanss$layer_code)
corr <- corr[unlist(correlation_groups(corr)), unlist(correlation_groups(corr))]
corr[corr==1] <- NA
plot_correlation(corr)
```

## Visualize QC

- 28\. Species outliers (geography): is the observation within three IQRs from the first & third quartile distance to the geographic centroid of this taxon?

```{r getacistu, cache=TRUE}
acistu <- occurrence("Acipenser sturio") # European sea sturgeon

acistu$qcnum <- qcflags(acistu$qc, c(28))
colors <- c("#ee3300", "#86b300")[acistu$qcnum + 1]
popup <- paste0(acistu$datasetName, "<br/>", acistu$catalogNumber, 
                "<br/><a href=\"http://www.iobis.org/explore/#/dataset/", 
                acistu$resourceID, "\">OBIS dataset page</a>")
```

## Visualize QC

```{r acistuleafletqc, cache=TRUE, message = FALSE}
library(leaflet)
leaflet() %>% addProviderTiles("CartoDB.Positron") %>%
  addCircleMarkers(
    popup = popup, radius = 3.5, weight = 0, fillColor = colors, fillOpacity = 1,
    lat = acistu$decimalLatitude, lng = acistu$decimalLongitude)
```

## Visualize QC

```{r acistuggplotqc, cache=TRUE}
acistu$qctxt <- c("Not ok", "Ok")[acistu$qcnum + 1]
ggplot() +
  geom_polygon(data=map_data("world"),aes(x=long,y=lat,group=group),fill="#dddddd")+
  geom_point(data=acistu, aes(x=decimalLongitude, y=decimalLatitude, color=qctxt))
```

## Visualize environmental data

```{r acistuenvplot, cache=TRUE}
# devtools::install_github('iobis/obistools')
env <- obistools::lookup_xy(acistu)
envacistu <- cbind(acistu[complete.cases(env),], env[complete.cases(env),])
ggplot(envacistu, aes(x=sssalinity, y=sstemperature, color=qctxt)) +
  geom_point()
```

## World + SST + MEOW + Abra alba

```{r mapdemo, cache=TRUE}
meow <- read_sf('data/MEOW/meow_ecos.shp')
sst <- raster('data/BO2_tempmean_ss_lonlat.tif')
extent <- extent(-87, 25, 24, 70)
sst <- raster::crop(sst, extent)
sst_df <- as.data.frame(as(sst, "SpatialPixelsDataFrame"))
colnames(sst_df) <- c("value", "x", "y")
map <- ggplot() +
  geom_tile(data=sst_df, aes(x=x, y=y, fill=value), alpha=0.8) +
  geom_sf(data=meow, fill="darkgreen", alpha=0.1) +
  geom_polygon(data=map_data("world"),aes(x=long,y=lat,group=group),fill="#dddddd") +
  geom_point(data=abra, aes(x=decimalLongitude, y=decimalLatitude)) +
  coord_sf(xlim = c(extent@xmin+10, extent@xmax-10), 
           ylim = c(extent@ymin+10, extent@ymax-10)) +
  scale_fill_distiller(palette = "Spectral") +
  labs(x="longitude", y="latitude", fill = "SST")
```

## World + SST + MEOW + Abra alba

```{r plotmapdemo, cache=TRUE, echo=FALSE}
map
```

## Exercises

- Plot thinned records and original records for e.g. Abra alba, zoom in on the European distribution
- Plot surface and benthic temperature against depth for Europe

